---
title: "01_clean-data"
author: "Francesco Grassi"
date: "2022-10-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Aim

This script cleans up the file containing raw data about deaths in the hometown.  

The file comes from: https://www.istat.it/storage/dati_mortalita/decessi-comunali-giornalieri-regioni(excel)-21-07-2022-5.zip  
The link downloads a .zip folder with an .xlsx file for each Italian region. Each file contains data regarding the dealy number of deaths in each town of the region in the period 01/01/2011 - 31/07/2022.  
The file of interest is "comuni_giornaliero_31luglio2022_Puglia.xlsx", which has been renamed in this project as "raw_all-region".

The file contains the following variables:

- REG: Code indicating Italian Region
- PROV: Code indicating Italian Province within Region
- NOME_REGIONE: Name of the Region
- NOME_PROVINCIA: Name of the Province
- NOME_COMUNE: Name of the city/town
- COD_PROVCOM: Code indicating the city/town
- TIPO_COMUNE: 1 if data is available for the entire assessment period; 2 if data from 2022 is missing
- CL_ETA: Age range of the deceased person
  + 0: 0 
  + 1: 1-4 
  + 2: 5-9
  + 3: 10-14 
  + 4: 15-19 
  + 5: 20-24 
  + 6: 25-29 
  + 7: 30-34 
  + 8: 35-39 
  + 9: 40-44 
  + 10: 45-49 
  + 11: 50-54 
  + 12: 55-59 
  + 13: 60-64 
  + 14: 65-69 
  + 15: 70-74 
  + 16: 75-79 
  + 17: 80-84 
  + 18: 85-89 
  + 19: 90-94 
  + 20: 95-99 
  + 21: 100+
- GE: day of death (MM/DD)
- Many columns with name structure "GENDER_YY", indicating number of deaths for a given GENDER ("M": Male; "F": Female, "T": Total) for the year YY (range: 2011-2022)

# Libraries
```{r}
library(tidyverse)
library(readxl)
library(lubridate)
```


# First cleaning

- Load raw data
- Select only the town of "Castelluccio dei Sauri"
- Get rid of all Region/Province names and codes, as well as "Total" death count columns
- Convert variable "GE" in date format "%d-%m"
- Replace "n.d." in death counts with NAs
- Convert death counts columns to numeric

```{r}

raw_data <- read_excel(path = "data/raw_all-region.xlsx", trim_ws = TRUE)

raw_data <- raw_data %>% 
  filter(NOME_COMUNE == "Castelluccio dei Sauri") %>% 
  select(!c(REG:TIPO_COMUNE, starts_with("T_"))) %>% 
  mutate(GE = format(as.Date(GE, format = "%m%d"), "%d-%m")) %>% 
  na_if("n.d.") %>% 
  mutate_at(vars(-CL_ETA, -GE), as.numeric)

```

# Deal with Dates

While column "GE" contains the day and month of decease, the year is stored in the variable name of the death counts columns. In order to obtain a proper date variable:

- Convert data.frame to long format, splitting death count variable names in "Gender" and "Year"
- Combine "GE" and "Year" into one variable and convert to date format

```{r}

raw_data_long <- raw_data %>% 
  pivot_longer(!c(CL_ETA, GE),
               names_to = c("Gender", "Year"),
               names_sep = "_",
               values_to = "Count") %>% 
  unite("Date", GE, Year, sep = "-") %>% 
  mutate(Date = dmy(Date))

```

# Final cleaning

- Remove rows with no death (0 or NAs in "Count")
- Replace "CL_ETA" with corresponding age range
- Rename variables
- Save final data.frame

```{r}

raw_data_long <- raw_data_long %>%
  filter(Count > 0) %>%
  mutate(CL_ETA = recode(CL_ETA,
    `0` = "0",
    `1` = "1-4",
    `2` = "5-9",
    `3` = "10-14",
    `4` = "15-19",
    `5` = "20-24",
    `6` = "25-29",
    `7` = "30-34",
    `8` = "35-39",
    `9` = "40-44",
    `10` = "45-49",
    `11` = "50-54",
    `12` = "55-59",
    `13` = "60-64",
    `14` = "65-69",
    `15` = "70-74",
    `16` = "75-79",
    `17` = "80-84",
    `18` = "85-89",
    `19` = "90-94",
    `20` = "95-99",
    `21` = "100+"
  )) %>% 
  rename(Age_Range = CL_ETA)
  
write_csv(raw_data_long, file = "data/clean-data.csv")

```


